Disease Knowledge Transfer across Alzheimer's Variants

### 2018-03-03 paper/paper.tex
final version for MICCAI ... my brain is dead

\begin{abstract}
The availability of multimodal biomarker measurements in large scale clinical data is fundamental for the application of current disease progression models of neurodegenerative disorders. Due to the lack of data and to the low sample size, disease progression modeling of rare dementias is currently very challenging. To overcome this problem, we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarker's progression models between Alzheimer's disease (AD) variants. We assume that different types of dementia affect overlapping brain regions, and thus present shared biomarker characteristics that can be transferred across diseases. We then implement this paradigm as a joint-disease generative model of biomarker progressions, disentangling disease-specific from  disease-agnostic biomarker relationships. We demonstrate DKT on independent datasets of 1) multimodal typical AD data (tAD), with large sample size and several biomarkers available over time, and 2) Posterior Cortical Atrophy (PCA) data, for which only MRI scans are available. DKT is able to predict, in tested PCA subjects, plausible population-level biomarkers for structural and molecular imaging biomarkers, for which no data were available. DKT may be a useful tool to analyse and understand rare forms of dementia for which multimodal data is not available or is limited. Moreover, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models.

@@

Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements (\cite{lorenzi2017disease,young2014data,donohue2014estimating}). These approaches mostly rely on the estimation of a latent time reparameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.  This time reparameterization can be used as a proxy of disease staging associated with an individual biomarker profile relatively to the global disease progression model.

% Flow: multimodal data needed for disease progression -> disease progression on rare dementias is challenging -> cannot apply model learned from other disease because of different spatial patterns -> but studies show at least some partial spatial overlap -> therefore transfer learning possible due to presence of "some overlap" -> benefits of transfer learning -> existing transfer learning literature
The availability of multimodal collections of biomarker's measurements across clinical groups is fundamental to the application of such disease progression models. For this reason, the application of disease progression modeling to rare dementia type is currently very challenging, due to the associated problem of missing biomarkers and low sample size.  Moreover, an average model of disease progression estimated from sporadic AD cases may not generalize to specific disease variants. For example, in Posterior Cortical Atrophy (PCA), posterior regions such as the occipital lobe and superior parietal regions have been shown to be affected early. While the spatial patterns are distinct between various forms of dementia, recent studies \cite{Ossenkoppele2014atrophy} have showed at least partial spatial-overlap between different clinical phenotypes, while others have further suggested that PCA and typical AD subjects lie on a continuum of phenotypical variation \cite{crutch2012posterior}. The presence of overlap in pathology patterns across different dementias suggests that it should be theoretically possible to perform transfer learning across the diseases. There are two key benefits to performing transfer learning across dementia variants: 1. biomarker evolutions can be estimated for rare dementias for which there is not enough data (unimodal datasets, few subjects, cross-sectional data only) and 2. adding extra information from other datasets can help with staging and biomarker trajectory estimation. While recent studies \cite{hon2017towards} performed transfer learning from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants, either for diagnosis classification or for estimating continuous biomarker trajectories.

@@

Fig. \ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of diseases. We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each one related to different brain regions (top row). Each dysfunctionality score is then modelled as the progression of several biomarkers within that same region, but having different modalities (bottom row). Each group of biomarkers in the bottom row will be called a \emph{functional unit}, because the correlations between biomarkers are related though common "function" in a disease agnostic way, since they are related to the same underlying brain region. Biomarker groupings into functional units are defined a-priori. We model the correlations within each unit using a Gaussian-Process (GP) based disease progression model (DPM) \cite{lorenzi2017disease}, which allows us to reconstruct unit-specific \emph{dysfunction} progression manifold (X axis), which can be used for staging subjects. Finally, we use the same GP model to express the progression within each disease (Figure 1, top) in terms of the dysfunction scores estimated within each functional unit. More precisely, the X-axis dysfunction scores from the functional units become Y-axis measurements in the disease specific models.

The model has a simple mathematical formulation. We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers in at various visits. We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the subject-spefic time shift. We futher denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its functional unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps each biomarker $k$ to a unique functional unit $l \in \Lambda$ and $\Lambda$ is the set of functional units. Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to functional unit $l \in \Lambda$ in the space of disease $d$. These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:

@@

We used 87 controls and 76 PCA subjects from our local dataset, which only had MRI scans. From TADPOLE/ADNI, out of the subjects that contained at least one FDG PET, AV45, AV1451 or DTI scan, we randomly selected a subset of 230 subjects, in order not to bias the numbers towards the typical AD disease.  Volumetric measures for each subject have been obtained using the Freesurfer software in both ADNI and our local dataset. For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI. For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe. For every lobe, we averaged the biomarker values for regions of interest within each lobe, regressed out age, gender, TIV and dataset (ADNI vs our local dataset) and then normalised the final biomarker values to lie within the [0,1] range.

@@

% Fig1: biomarker traj. over dysfunction scores in one functional unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Fig. \ref{fig:pcaTadDisSpace}A shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data. The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range. The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. Fig. \ref{fig:pcaTadDisSpace} also shows the progression of dysfunctionality scores over the disease stage for (B) typical AD and (C) PCA. In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior} (J.C. Baron, Neuroimage, 2001). In Fig. \ref{fig:PCAtrajByModality}, we plot the inferred biomarker trajectories for PCA directly across the disease progression. We do this for five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451. The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, with the exception of FDG. This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in artificially more scaling to trajectories that contain less disease signal. We also notice two non-monotonic trajectories in the AV45 plot, which are due to bad starting points within parameter optimisation combined with low disease signal within these biomarkers.

@@

% data fitting within one unit
% \begin{figure}
% \centering
% \includegraphics[width=0.8\textwidth, trim=90 0 110 0, clip]{figures/unit1_allTraj_tad-drcTinyPen1_JMD.png} 
% \caption{Estimated biomarker trajectories in the Occipital Unit. Subject data from ADNI and our local cohort are also shown. The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject. Red lines represent samples from the trajectory posterior. The Y-axis measures biomarker values (normalised).}
% \label{fig:occipUnit}
% \end{figure}

@@

\begin{picture}(5,5)
\put(0,20){\textbf{\Large{B}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% typical AD\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
% \caption{C}
\end{subfigure}
\begin{picture}(5,5)
\put(0,20){\textbf{\Large{C}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% PCA\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
% \caption{PCA}
\end{subfigure}
\caption{(A) Estimated biomarker trajectories in the Occipital Unit. Subject data from ADNI and our local cohort are also shown. The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject. Red lines represent samples from the trajectory posterior. The Y-axis measures biomarker values (normalised). (B-C) Progression of dysfunctionality scores for (B) typical AD and (C) PCA.}
\label{fig:pcaTadDisSpace}
\end{figure}

@@

We performed validation using a separate test set of 20 DTI scans from controls and PCA patients from our local cohort. Fig. \ref{fig:DTIvalid} shows the estimated DTI biomarker trajectories plotted directly against the disease progression time-shifts, along with the DTI data from the test set. The time shift of each subject was estimated only based on their MRI scans. The inferred trajectories in PCA show good agreement with the data in the cingulate, hippocampus and temporal lobes, but less agreement in the other regions. This might be because the dataset that was used to train the model had better disease signal in DTI regions more traditionally associated with typical AD.

@@

% how to address aforementioned limitations + future work
There are several potential avenues for further research. During training, we can account for missing disease signal in some brain regions by adding other types of dementias that affect other regions. The estimated PCA trajectories also need to be thoroughly validated using data from other modalities. The model formulation can also be naturally extended to include subject-specific effects.

@@

\bibitem{young2014data}
Young, A.L., Oxtoby, N.P., Daga, P., Cash, D.M., Fox, N.C., Ourselin, S., Schott, J.M. and Alexander, D.C., 2014. A data-driven model of biomarker changes in sporadic Alzheimer's disease. Brain, 137(9), pp.2564-2577.

\bibitem{donohue2014estimating}
Donohue, M.C., Jacqmin-Gadda, H., Le Goff, M., Thomas, R.G., Raman, R., Gamst, A.C., Beckett, L.A., Jack, C.R., Weiner, M.W., Dartigues, J.F. and Aisen, P.S., 2014. Estimating long-term multivariate progression from short-term data. Alzheimer's \& Dementia, 10(5), pp.S400-S410.

@@

>>>>

\begin{abstract}
The availability of multimodal biomarker measurements in large scale clinical data is fundamental for the application of current disease progression models of neurodegenerative disorders. Due to the lack of data and to the low sample size, disease progression modeling of rare dementias is currently very challenging. To overcome this problem, we introduce Disease Knowledge Transfer (DKT), a technique for transferring biomarkers' progression models between Alzheimer's disease (AD) variants. We assume that different types of dementia affect overlapping brain regions, and thus present shared biomarker characteristics that can be transferred across diseases. We then implement this paradigm as a joint-disease generative model of biomarker progressions, disentangling disease-specific from  disease-agnostic biomarker relationships. We demonstrate DKT on independent datasets of 1) multimodal typical AD data (tAD), with large sample size and several biomarkers available over time, and 2) Posterior Cortical Atrophy (PCA) data, for which only MRI scans are available. DKT is able to predict, in tested PCA subjects, plausible population-level biomarkers for structural and molecular imaging biomarkers, for which no data were available. DKT may be a useful tool to analyse and understand rare forms of dementia for which multimodal data is not available or is limited. Moreover, by leveraging data from multiple diseases, DKT also has the potential to provide more accurate disease staging compared to traditional disease progression models.

@@

Several data-driven methods have been proposed in order to reconstruct group-wise long term biomarker progressions from collections of short term individual biomarker measurements \cite{lorenzi2017disease,oxtoby2018}. These approaches mostly rely on the estimation of a latent time re-parameterization associated with each individual, often encoded by a translation (time shift) of the subject's measurements over the temporal axis.  This time re-parameterization can be used as a proxy of disease staging associated with an individual biomarker profile relatively to the global disease progression model.

% Flow: multimodal data needed for disease progression -> disease progression on rare dementias is challenging -> cannot apply model learned from other disease because of different spatial patterns -> but studies show at least some partial spatial overlap -> therefore transfer learning possible due to presence of "some overlap" -> benefits of transfer learning -> existing transfer learning literature
The availability of multimodal collections of biomarkers' measurements across clinical groups is fundamental to the application of such disease progression models. For this reason, the application of disease progression modeling to rare dementia type is currently very challenging, due to the associated problem of missing biomarkers and low sample size.  Moreover, an average model of disease progression estimated from sporadic AD cases may not generalize to specific disease variants. For example, in Posterior Cortical Atrophy (PCA), posterior regions such as the occipital lobe and superior parietal regions have been shown to be affected early. While the spatial patterns are distinct between various forms of dementia, recent studies \cite{Ossenkoppele2014atrophy} have showed at least partial spatial-overlap between different clinical phenotypes, while others have further suggested that PCA and typical AD subjects lie on a continuum of phenotypical variation \cite{crutch2012posterior}. The presence of overlap in pathology patterns across different dementias suggests that it should be theoretically possible to perform transfer learning across the diseases. There are two key benefits to performing transfer learning across dementia variants: 1. biomarker evolutions can be estimated for rare dementias for which there is not enough data (unimodal datasets, few subjects, cross-sectional data only) and 2. adding extra information from other datasets can help with staging and biomarker trajectory estimation. While recent studies \cite{hon2017towards} performed transfer learning from generic image datasets to Alzheimer's disease, we are not aware of any studies that have tried to transfer knowledge across different types of Alzheimer's variants, either for diagnosis classification or for estimating continuous biomarker trajectories.

@@

Fig. \ref{fig:diagram} shows the overall diagram of our proposed framework for joint modelling of diseases. We assume that each disease can be modelled as the evolution of abstract dysfunctionality scores, each one related to different brain regions (top row). Each dysfunctionality score is then modelled as the progression of several biomarkers within that same region, but having different modalities (bottom row). Each group of biomarkers in the bottom row will be called a \emph{functional unit}, because the correlations between biomarkers are related though common "function" in a disease agnostic way, since they are related to the same underlying brain region. Biomarker groupings into functional units are defined a-priori. We model the correlations within each unit using a Gaussian-Process (GP) based disease progression model (DPM) \cite{lorenzi2017disease}, which allows us to reconstruct unit-specific \emph{dysfunction} progression manifolds (bottom row, X axis), which can be used for staging subjects. Finally, we use the same GP model to express the progression within each disease (Figure 1, top) in terms of the dysfunction scores estimated within each functional unit. More precisely, the X-axis dysfunction scores from the functional units become Y-axis measurements in the disease specific models.

The model has a simple mathematical formulation. We assume a set of given biomarkers measurements $Y = [y_{ijk} | (i,j,k) \in \Omega]$ for subject $i$ at visit $j$ in biomarker $k$, where $\Omega$ is defined as the set of available biomarker measurements, since subjects can have missing biomarkers at various visits. We assume that each subject $i$ at each visit $j$ has an underlying disease stage $s_{ij} = \beta_i + m_{ij}$, where $m_{ij}$ represents the months since baseline visit for subject $i$ at visit $j$ and $\beta_i$ represents the time shift of subject $i$. We further denote by $\theta_k$ the parameters used to represent the trajectory for biomarker $k \in K$ within its functional unit $\psi(k)$, where $\psi$: \{1, ..., K\} $ \rightarrow \Lambda$ is a function that maps each biomarker $k$ to a unique functional unit $l \in \Lambda$, where $\Lambda$ is the set of functional units. Moreover, we denote by $\lambda_d^l$ the parameters for the trajectory of the dysfunction score corresponding to functional unit $l \in \Lambda$ in the space of disease $d$. These definitions allow us to formulate the likelihood for a single measurement $y_{ijk}$ as follows:

@@

We used 87 controls and 76 PCA subjects from our local dataset, which only had MRI scans. From TADPOLE/ADNI, out of the subjects that contained at least one FDG PET, AV45, AV1451 or DTI scan, we randomly selected a subset of 230 subjects, in order not to bias the numbers towards the typical AD disease.  Volumetric measures for each subject have been obtained using the Freesurfer software in both ADNI and our local dataset. For FDG, AV45 and AV1451 PET, we used already extracted SUVR measures from ADNI. For DTI, we used fractional anisotropy (FA) measures from white-matter regions adjacent to each lobe. For every lobe, we averaged the biomarker values for regions of interest within each lobe, regressed out age, gender, total intracranial volume (TIV) and dataset (ADNI vs our local dataset) and then normalised the final biomarker values to lie within the [0,1] range.

@@

% Fig1: biomarker traj. over dysfunction scores in one functional unit -> Fig2: dysfunction trajectories over disease stage in the two disease models -> Fig3: inferred biomarker trajectories "directly" over disease stage  in PCA
Fig. \ref{fig:pcaTadDisSpace}A shows the estimated biomarker trajectories within the \emph{occipital unit} plotted over the dysfunction scores, along with samples from the model posterior and aligned subject data. The X-axis shows the dysfunctionality scores within the occipital unit, which represent estimated time-shifts, in months, from an arbitrary reference X=0, while the Y-axis shows biomarker values normalised to [0,1] range. The model shows a good data fit, and we can observe most PCA subjects having abnormal occipital volumes, thus leading to high occipital dysfunctionality scores, in line with the current understanding of PCA as affecting posterior regions \cite{crutch2012posterior}. Fig. \ref{fig:pcaTadDisSpace} also shows the progression of dysfunctionality scores over the disease stage for (B) typical AD and (C) PCA. In typical AD, we see that temporal dysfunction becomes abnormal earliest, while PCA shows early parietal dysfunction, again in line with previous findings in the literature \cite{crutch2012posterior,Baron2001}. In Fig. \ref{fig:PCAtrajByModality}, we plot the inferred biomarker trajectories for PCA directly across the disease progression. We do this for five different modalities: MRI Volumes, DTI, FDG, AV45 and AV1451. The results again recapitulate known patterns in PCA, where trajectories from posterior regions become abnormal first in MRI, DTI and AV1451, with the exception of FDG. This might be due to the normalisation of the FDG trajectories on the Y-axis to the [0,1] range, which results in artificially more scaling to trajectories with less disease signal. We also notice two non-monotonic trajectories in the AV45 plot, which are due to bad starting points within parameter optimisation combined with low disease signal within these biomarkers.

@@

@@

\begin{picture}(5,5)
\put(0,20){\textbf{\Large{B}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% typical AD\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/tAD_trajSameSpace_tad-drcTinyPen1_JMD.png} 
% \caption{C}
\end{subfigure}
\begin{picture}(5,5)
\put(0,20){\textbf{\Large{C}}}
\end{picture}
\begin{subfigure}{0.47\textwidth}
\centering
% PCA\\
\includegraphics[width=0.8\textwidth, trim=0 0 0 20, clip]{figures/PCA_trajSameSpace_tad-drcTinyPen1_JMD.png} 
% \caption{PCA}
\end{subfigure}
\caption{(A) Estimated biomarker trajectories in the occipital functional unit. Subject data from ADNI and our local cohort are also shown. The X-axis, defined as the occipital dysfunctionality score, represents the time-shifts (in months) of each subject. Red lines represent samples from the trajectory posterior. The Y-axis measures biomarker values (normalised). (B-C) Progression of dysfunctionality scores for (B) typical AD and (C) PCA.}
\label{fig:pcaTadDisSpace}
\end{figure}

@@

We performed validation using a separate test set of 20 DTI scans from controls and PCA patients from our local cohort. Fig. \ref{fig:DTIvalid} shows the estimated DTI biomarker trajectories plotted directly against the disease progression time-shifts, along with the DTI data from the test set. The time shift of each subject was estimated only based on their MRI scans. The inferred trajectories in PCA show good agreement with the data in the cingulate, hippocampus and temporal lobes, but less agreement in the other regions. This might be because the tAD DTI dataset that was used to train the model had better disease signal in regions traditionally associated with typical AD.

@@

% how to address aforementioned limitations + future work
There are several potential avenues for further research. During training, we can account for missing disease signal in some brain regions by adding other types of dementias that affect other regions. The estimated PCA trajectories also need to be thoroughly validated using data from other modalities. The model formulation also extends naturally to include subject-specific effects.

@@

\bibitem{oxtoby2018}
Oxtoby, N.P., Young, A.L., Cash, D.M., Benzinger, T.L., Fagan, A.M., Morris, J.C., Bateman, R.J., Fox, N.C., Schott, J.M. and Alexander, D.C., 2018. Data-driven models of dominantly-inherited Alzheimer's disease progression. bioRxiv, p.250654.

@@

\bibitem{Baron2001}
Baron, J.C., Ch\'{e}telat, G., Desgranges, B., Perchey, G., Landeau, B., {de la Sayette}, V. and Eustache, F., 2001. NeuroImage 14(2), pp.298-309.
